---
title: "Margaret RNA DESeq Skeleton"
output: html_document
date: "2025-07-31"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Preface
The goal of this script is to identify genes differentially expressed between purple and white flowers in I. hederacea and to map them to the anthocyanin pathway using KEGG pathways. 

```{r library_loading}
library(DESeq2)
library(ggplot2)
library(tidyr)
library(stringr)
library(dplyr)
library(KEGGgraph)
library(readODS)
library(igraph)
library(ComplexHeatmap)
library(circlize)
library(patchwork)
```

## Initiating Variables
```{r var_init}
genecountdir = "../processed_data/20250730_star_align_out/" #directory containing the read counts from STAR
strandcolumn = 4 #whether to use unstranded (2), forward stranded (3), or reverse stranded (4) counts
grouplist1 = c("white", "white", "purple", "purple", "white", "purple", "white", "purple") #levels of factor of interest in alphabetical order of files (i.e., the same order as list.files output)
group1_order = c("white", "purple") #order of levels you want for group1 so that control/low group is always the same
test_type = "LRT" #type of test for DESeq analysis. Wald is preferred if you're just doing your model against a full null. LRT is useful if you're comparing between a full model and a simplified model
alpha_cutoff = 0.05 #alpha cut-off for multiple hypothesis correction of DESeq2 results
color_palette1 = c("grey80", "#c371e2") #color palette for plotting. colors will correspond to the order of the levels in group1_order
genome_manifest = "../manual_manifests/20250806_Ihederaceae_araport11_diamondblastp_moresensitive.csv" #path to file containing DIAMOND/BLAST search of reference genome hits against arabidopsis database. Note, we may change this to tomato later in the script because that is a closer species with more variation in flower color and a lot of the genes in the anthocyanin pathway are within the same gene family. So, something as distantly related as arabidopsis thaliana may not be a good reference. 
```

## Dataframe Management
```{r data_management}
### Build gene counts dataframe ###
filelist = list.files(genecountdir, pattern = "*ReadsPerGene.out.tab")

#initialize counts dataframe by reading in first file
counts = read.table(paste(genecountdir, filelist[1], sep = ""))
row.names(counts) = counts$V1
counts = counts[,c(1, strandcolumn)]
colnames(counts)[2] = filelist[1]
counts[,3:(length(filelist)+1)] = NA
counts$V1 = NULL

for (i in 2:length(filelist)) {
  tempdf = read.table(paste(genecountdir, filelist[i], sep = ""))
  counts[,i] = tempdf[,strandcolumn]
  colnames(counts)[i] = filelist[i]
}

### Build metadata dataframe (skip if you already have this) ###
meta_df = data.frame(
  sample_name = filelist,
  group1 = grouplist1
)

row.names(meta_df) = meta_df$sample_name

### Prepare dataframes for input into DESeq2 ###
#put metadata in order of column names from gene counts matrix
all(rownames(meta_df) %in% colnames(counts)) #check all samples are in counts
all(rownames(meta_df) == colnames(counts)) #check samples are in order
counts = counts[, rownames(meta_df)] #put data in same order if the first is true but the second is false

#remove STAR info about unmapped, multimapped etc. reads
counts = counts[5:dim(counts)[1],]

#order factor levels for DESeq to know which is the reference group
meta_df$group1 = factor(meta_df$group1, levels = group1_order)
```

## DESeq2
```{r deseq2}
#load data in to DESeq object
dds = DESeqDataSetFromMatrix(countData = counts,
                                 colData = meta_df,
                                 design = ~group1
                                 )

#remove low count reads (optional, but I'm doing it)
keep = rowSums(counts(dds)) >= 10
dds = dds[keep,]

#perform differential expression analysis
if (test_type == "Wald") {
  dds_analysis = DESeq(dds, test = test_type) 
} else {
    dds_analysis = DESeq(dds, test = test_type, reduced = ~1)   
}

res = results(dds_analysis, alpha = alpha_cutoff)

mcols(res, use.names = TRUE) #prints descriptions of columns in results df
print(sum(res$padj < alpha_cutoff, na.rm = TRUE)) #prints number of significant hits

#get significant results
res = res[!is.na(res$padj),] #DESeq2 will introduce NAs when adjusting for p-values because, I think, these are genes that are too weakly expressed and/or are constant (e.g., all 0s) across all samples.
sig_genes = res[res$padj < alpha_cutoff,]

#print out list of significant genes
print(rownames(sig_genes))

#read in genome manifest to add information about significant hits
refseq_hits = read.csv(genome_manifest)

#add refseq hit info
sig_genes_descriptive = merge(as.data.frame(sig_genes), refseq_hits, by.x = 0, by.y = "ref_gene_id", all.x = TRUE)
sig_genes_descriptive$enrichment = ifelse(sig_genes_descriptive$log2FoldChange < 0, group1_order[1], group1_order[2])

#export results
write.table(sig_genes_descriptive, "../processed_data/20250804_margaret_whitevpurple_significantDESeqgenes_LRT_araport11.tsv", row.names = FALSE, quote = FALSE, sep = "\t")
```

## Quick plotting of DESeq2 results

```{r}
res$enrichment = ifelse(res$log2FoldChange < 0, group1_order[1], group1_order[2])

colorman1 = c(color_palette1)
names(colorman1) = group1_order

#Volcano Plot
ggplot(as.data.frame(res), aes(x = log2FoldChange, y = -log(padj))) +
  geom_point(aes(color = enrichment, alpha = padj < 0.05)) +
  geom_hline(yintercept = -log(alpha_cutoff)) +
  theme_classic() +
  scale_color_manual(values = colorman1) +
  geom_text(sig_genes, mapping = aes(x = log2FoldChange, y = -log(padj), label = rownames(sig_genes)))

```
```{r}
plotMA(res)
```
```{r}
plotCounts(dds, gene = which.min(res$padj), intgroup = "group1")
```
## Selecting anthocyanin genes

```{r}
#read in .keg hierarchy file
kegg_file = readLines("../raw_data/ath00001.keg")

#only keep lines that start with D. these are the lowest level of the hierarchy that contains matches between A. thaliana genes and KEGG info
kegg_genemap = c()
for (i in 1:length(kegg_file)) {
  if (substr(kegg_file[i], 1, 1) == "D") {
    kegg_genemap = append(kegg_genemap, kegg_file[i])
  }
}

#some text formatting to make this gene map into a dataframe compatible format
kegg_formatting = sapply(kegg_genemap, function(x) gsub("D      ", "", x))
kegg_formatting = unname(kegg_formatting)

at_kegg = rep(NA, length(kegg_formatting))
kegg_kegg = rep(NA, length(kegg_formatting))
for (i in 1:length(kegg_formatting)) {
  holder = str_split(kegg_formatting[i], "\t")
  at_kegg[i] = holder[[1]][1]
  kegg_kegg[i] = holder[[1]][2]
}

kegg_df = data.frame(at_kegg, kegg_kegg)
kegg_df$at_gene_id = sapply(kegg_df$at_kegg, function(x) substr(x, 1, 9))

at_gene_symbol = rep(NA, dim(kegg_df)[1])
at_gene_long = rep(NA, dim(kegg_df)[1])
for (i in 1:dim(kegg_df)[1]) {
  holder = strsplit(kegg_df$at_kegg[i], "; ", fixed = TRUE)
  if (is.na(holder[[1]][2])) {
    at_gene_symbol[i] = holder[[1]][1]
    at_gene_long[i] = holder[[1]][1]
  }
  else {
    at_gene_symbol[i] = substr(holder[[1]][1], 11, str_length(holder[[1]][1]))
    at_gene_long[i] = holder[[1]][2]
  }
}
kegg_df$at_gene_symbol = at_gene_symbol
kegg_df$at_gene_long = at_gene_long

kegg_df$kegg_id = substr(kegg_df$kegg_kegg, 1, 6)

kegg_symbol = rep(NA, dim(kegg_df)[1])
kegg_long = rep(NA, dim(kegg_df)[1])
for (i in 1:dim(kegg_df)[1]) {
  holder = strsplit(kegg_df$kegg_kegg[i], "; ", fixed = TRUE)
  if (is.na(holder[[1]][2])) {
    kegg_symbol[i] = holder[[1]][1]
    kegg_long[i] = holder[[1]][1]
  }
  else {
    kegg_symbol[i] = substr(holder[[1]][1], 8, str_length(holder[[1]][1]))
    kegg_long[i] = holder[[1]][2]
  }
}
kegg_df$kegg_symbol = kegg_symbol
kegg_df$kegg_long = kegg_long

kegg_df$ec_id = sapply(kegg_df$kegg_long, function(x) str_extract(x, "(?<=\\[EC\\:)[^\\]]+(?=\\])"))

kegg_df$at_kegg = NULL
kegg_df$kegg_kegg = NULL

write.table(kegg_df, "../processed_data/20250807_arabidopsis_kegg_map.tsv", sep = "\t", quote = FALSE, row.names = FALSE)
```

## Gene Expression Heatmap
```{r}
samp_exp = rlog(dds, blind = TRUE)

samp_exp_df = as.data.frame(assay(samp_exp))

#create map of anthocyanin pathway genes to ipomoea arabidopsis hits
antho_genes = read_ods("../manual_manifests/manual_network_tables.ods", sheet = 4) #some of the pathway genes are missing in arabidopsis, so we will have to manually add from whole nr search
refseq_hits$repo_gene_id = sapply(refseq_hits$repo_accession, function(x) strsplit(x, ".", fixed = TRUE)[[1]][1])
ref_antho = refseq_hits[refseq_hits$repo_gene_id %in% antho_genes$arabidopsis_id,]

#only keep entries where the alignment coverage of the ipomoea gene is more than 70%
ref_antho = ref_antho[ref_antho$ref_coverage >= .7,]

#get top representative for each gene
ref_antho = ref_antho %>%
  group_by(ref_gene_id) %>%
  filter(evalue == min(evalue)) %>%
  filter(row_number() == 1)

ref_antho = merge(ref_antho, antho_genes, by.x = "repo_gene_id", by.y = "arabidopsis_id")

#let's add g2401 to the list which, in nil, is closer to UF3GT and 3'GT. It's hard to say which is which since blasts results are really similar
antho_exp = samp_exp_df[row.names(samp_exp_df) %in% ref_antho$ref_gene_id,]
antho_exp[dim(antho_exp)[1]+1,] = samp_exp_df[rownames(samp_exp_df) == "g2401",]

#reorder the columns so white is on one side and purple on the other
antho_exp = antho_exp[,c(1,2,5,7,3,4,6,8)]

antho_exp$ref_gene_id = row.names(antho_exp)
antho_exp = merge(antho_exp, ref_antho, by = "ref_gene_id", all.x = TRUE)

simple_exp = as.matrix(antho_exp[,2:9])
rownames(simple_exp) = antho_exp$ref_gene_id

row_map = antho_exp$gene_symbol_short
names(row_map) = antho_exp$ref_gene_id
col_fun = colorRamp2(c(0, max(simple_exp)/2, max(simple_exp)), c("#3B6BC4", "white", "#C4943B"))
Heatmap(simple_exp, col = col_fun, cluster_rows = FALSE, cluster_columns = FALSE,
        right_annotation = rowAnnotation(antho_gene = antho_exp$gene_symbol_short))
```

```{r}
#plotting the log fold change expression of anthocyanin pathway genes
antho_fc = as.data.frame(res)
antho_fc$ref_gene_id = row.names(antho_fc)
antho_fc = antho_fc[antho_fc$ref_gene_id %in% antho_exp$ref_gene_id,]

antho_fc = merge(antho_fc, ref_antho, by = "ref_gene_id", all.x = TRUE)
antho_fc[antho_fc$ref_gene_id == "g2401", "gene_symbol_short"] = "g2401"

#change level order of gene_symbol_short to roughly match the order in the pathway
antho_fc$gene_symbol_short = factor(antho_fc$gene_symbol_short, levels = c(
  "CHS", "CHI", "F3H", "F3’H", "DFR", "FLS", "ANS", "UF3GT", "5GT", "g2401"
))

ggplot(antho_fc, aes(x = gene_symbol_short, y = log2FoldChange, fill = gene_symbol_short, alpha = -log(padj))) +
  geom_col(position = position_dodge2(preserve = "single")) +
  theme_minimal() +
  geom_hline(yintercept = 0)
```

```{r}
antho_fc$gene_symbol_short = factor(antho_fc$gene_symbol_short, levels = rev(levels(antho_fc$gene_symbol_short)))

ggplot(antho_fc, aes(x = log2FoldChange, y = gene_symbol_short)) +
  geom_point(color = "#c371e2", alpha = ifelse(antho_fc$padj <= 0.05, 1, .3), size = 3) +
  geom_errorbarh(aes(xmax = log2FoldChange + lfcSE, xmin = log2FoldChange - lfcSE), color = "#c371e2", alpha = ifelse(antho_fc$padj <= 0.05, 1, .3), height = 0) +
  theme_minimal() +
  geom_vline(xintercept = 0) +
  theme(legend.position = "none") +
  xlab("log2(Fold Change)") +
  ylab("Gene")
```


## Get Homologs of Anthocyanin Genes Missing in Arabidopsis by Matching to KEGG Tomato Hits

The missing genes are: F3'5'H (EC:1.14.14.81; Tomato: 100736504, Tomato Uniprot: D3W9H7) and 3'GT (EC:2.4.1.238; Gentiana Triflora: BAC54092, Gentiana Uniprot: Q8H0F2). Also, according to PANTHER search, g2401 is most likely BZ1 (Anthocyanidin 3-O-glucosyltransferase, EC:2.4.1.115) which is missing in Arabidopsis so let's grab the tomato version (101260093, Uniprot: A0A3Q7IMM2).

```{r}
#read in DIAMOND blastp output of search for the missing genes in Ipomoea
f3p5ph_diamond = read.table("../manual_manifests/20250826_tomato_f35h_uniprot-d3w9h7_diamondblastp_moresensitive.txt", header = FALSE)
gen3pgt_diamond = read.table("../manual_manifests/20250826_gentiana_3pGT_uniprot-q8h0f2_diamondblastp_moresensitive.txt", header = FALSE)
bz1_diamond = read.table("../manual_manifests/20250826_tomato_bz1_uniprot-A0A3Q7IMM2_diamondblastp_moresensitive.txt", header = FALSE) #g2401 is the top hit here so pretty confident that's who it is

#get gene names of the top 5 hits to make comparable with the full genome search
f3p5ph_diamond = f3p5ph_diamond[1:5,]
gen3pgt_diamond = gen3pgt_diamond[1:5,]
bz1_diamond = bz1_diamond[1:5,]

f3p5ph_diamond$antho_gene = "F3'5'H"
gen3pgt_diamond$antho_gene = "3'GT"
bz1_diamond$antho_gene = "BZ1"

missing_df = rbind(f3p5ph_diamond, gen3pgt_diamond, bz1_diamond)
missing_df = missing_df[,6:dim(missing_df)[2]]
missing_df$repo_gene_id = sapply(missing_df$V6, function(x) strsplit(x, ".", fixed = TRUE)[[1]][1])
missing_df$V6 = NULL

missing_fc = as.data.frame(res)
missing_fc$ref_gene_id = row.names(missing_fc)
missing_fc = missing_fc[missing_fc$ref_gene_id %in% missing_df$repo_gene_id,]


missing_fc = merge(missing_fc, missing_df, by.x = "ref_gene_id", by.y = "repo_gene_id", all.x = TRUE)
colnames(missing_fc)[dim(missing_fc)[2]] = "gene_symbol_short"
colnames(missing_fc)[1] = "ref_gene_id"

#add this to the previous dataframe and remove the previous entry for g2401
antho_temp = antho_fc
antho_temp = antho_temp[antho_temp$ref_gene_id != "g2401",]
antho_temp = antho_temp[, colnames(antho_temp) %in% colnames(missing_fc)]

fullpath_fc = rbind(antho_temp, missing_fc)
fullpath_fc$gene_symbol_short = factor(fullpath_fc$gene_symbol_short, levels = rev(c(
  "CHS", "CHI", "F3H", "F3’H", "DFR", "FLS", "ANS", "F3'5'H",  "BZ1", "UF3GT", "5GT", "3'GT"
)))


ggplot(fullpath_fc, aes(x = log2FoldChange, y = gene_symbol_short)) +
  geom_point(color = "#c371e2", alpha = ifelse(fullpath_fc$padj <= 0.05, 1, .3), size = 3) +
  geom_errorbarh(aes(xmax = log2FoldChange + lfcSE, xmin = log2FoldChange - lfcSE), color = "#c371e2", alpha = ifelse(fullpath_fc$padj <= 0.05, 1, .3), height = 0) +
  theme_minimal() +
  annotate("text", x = fullpath_fc[fullpath_fc$ref_gene_id == "g2401", "log2FoldChange"], y = "BZ1", label = "g2401", vjust = -.7) +
  geom_vline(xintercept = 0) +
  theme(legend.position = "none") +
  xlab("log2(Fold Change)") +
  ylab("Gene")
```

## Exclusively only using tomato KEGGs

Because a lot of these genes are part of the same family, we may want to only use hits from a closely related species with a good annotation. This will be tomato. I am manually searching KEGG for these sequences from Solanum lycopersicum unless indicated otherwise in the species column of the input dataframe.

```{r}
#read in DIAMOND BLASTp results of tomato anthocyanin genes
tomato_kegg_blast = read.table("../manual_manifests/20250910_tomato_anthokegg_diamondblastp_moresensitive.txt", header = FALSE)

#filter for max top 3 hits per search
tom_filt = tomato_kegg_blast %>%
  group_by(V1) %>%
  slice_min(V3, n = 3)

tom_filt$antho_gene = sapply(strsplit(tom_filt$V1, "|", fixed = TRUE), function(x) x[1])
tom_filt$gene_id = sapply(strsplit(tom_filt$V2, ".", fixed = TRUE), function(x) x[1])

tom_filt = merge(tom_filt, as.data.frame(res), by.x = "gene_id", by.y = 0, all.x = TRUE)

tom_filt$antho_gene = factor(tom_filt$antho_gene, levels = rev(c(
  "CHS", "CHI", "F3H", "F3'H", "DFR", "FLS", "ANS", "F3'5'H",  "BZ1", "UF3GT"
)))


ggplot(tom_filt, aes(x = log2FoldChange, y = antho_gene)) +
  geom_point(color = ifelse(tom_filt$padj <= 0.05, "#c371e2", "grey"), size = 3, alpha = ifelse(tom_filt$padj <= 0.05, 1, .3)) +
  geom_errorbarh(aes(xmax = log2FoldChange + lfcSE, xmin = log2FoldChange - lfcSE), color = ifelse(tom_filt$padj <= 0.05, "#c371e2", "grey"), alpha = ifelse(tom_filt$padj <= 0.05, 1, .3), height = 0) +
  theme_minimal() +
  annotate("text", x = tom_filt[tom_filt$gene_id == "g2401", "log2FoldChange"], y = "BZ1", label = "g2401", vjust = -.7) +
  geom_vline(xintercept = 0) +
  theme(legend.position = "none") +
  xlab("log2(Fold Change)") +
  ylab("Gene")
```
## Base Image for Manuscript
We will build a two part plot using patchwork. We will then add svgs of white and purple flowers in Inkscape and make any small text edits in Inkscape too.

```{r}
purple_flower = "#6d6aa8ff"
color_palette2 = c("grey80", purple_flower)
colorman2 = c(color_palette2)
names(colorman2) = group1_order

(p1 = ggplot(as.data.frame(res), aes(x = log2FoldChange, y = -log(padj))) +
  geom_hline(yintercept = -log(alpha_cutoff), linetype = "dashed", color = "grey40") +
  geom_hline(yintercept = 0, color = "grey40") +
  geom_vline(xintercept = 0, color = "grey40") +
  geom_point(aes(color = enrichment, alpha = padj < 0.05)) +
  scale_color_manual(values = colorman2) +
  annotate("text", x = sig_genes[rownames(sig_genes) == "g2401", "log2FoldChange"], y = -log(sig_genes[rownames(sig_genes) == "g2401", "padj"]), label = "g2401", vjust = 1.5) +
  theme_minimal() +
   theme(legend.position = "none")
)

(p2 = ggplot(tom_filt, aes(x = log2FoldChange, y = antho_gene)) +
  geom_vline(xintercept = 0, color = "grey40") +
  geom_point(color = ifelse(tom_filt$padj <= 0.05, purple_flower, "grey"), size = 3, alpha = ifelse(tom_filt$padj <= 0.05, 1, .3)) +
  geom_errorbarh(aes(xmax = log2FoldChange + lfcSE, xmin = log2FoldChange - lfcSE), color = ifelse(tom_filt$padj <= 0.05, purple_flower, "grey"), alpha = ifelse(tom_filt$padj <= 0.05, 1, .3), height = 0) +
  theme_minimal() +
  annotate("text", x = tom_filt[tom_filt$gene_id == "g2401", "log2FoldChange"], y = "BZ1", label = "g2401", vjust = -.7) +
  theme(legend.position = "none") +
  xlab("log2(Fold Change)") +
  ylab("")
)
p1+p2

svg("../base_images/GOAL2-Transcriptome.svg", width = 7, height = 3.5)
p1+p2
dev.off()
```

